Astrophysical calibration of gravitational-wave detectors

### 2015-10-09 docs/cosmic_cal.tex
Some final changes

\begin{abstract}
  We present an investigation assessing the validity of the 
of gravitational wave detector calibration through the use of gamma-ray
bursts as standard
  sirens. Such signals, as measured via gravitational wave
  observations provide an estimated luminosity distance which is
  subject to uncertainties in the calibration of the data.  If a host
  galaxy is identified for a given source then its redshift can be
  combined with current knowledge of the cosmological
  parameters yielding the true luminosity distance.  This will then allow
  a direct comparison with the estimated value and discrepancies can
  then be used to correct the original calibration.~\MP{Update with results.
End with a statement comparing to existing method but stressing its independent.}
\end{abstract}

\pacs{04.80.Nn, 95.55.Ym}
\preprint{LIGO-P15XXXXX}

@@

% BNS systems are GRBs
It is considered likely that the merger of \ac{BNS} and/or \ac{NSBH} systems,
in addition to emitting detectable \acp{GW}, is also the mechanism for
producing \acp{sGRB} \cite{1992ApJ...395L..83N}. In this scenario these events
produce tightly beamed \ac{EM} emission parallel to the orbital angular
momentum vector of the \ac{BNS} system. Additional evidence for the coincidence
of \acp{sGRB} with \ac{GW} events are the estimated astrophysical rates of both
phenomena. Observations of \acp{sGRB} give an inferred rate of \ac{sGRB}
producing \ac{CBC} mergers of $8\times 10^{-9}$--$1.1\times
10^{-6}$\,Mpc$^{-3}$\,yr$^{-1}$~\cite{2012MNRAS.425.2668C} which can be
compared to independently obtained estimates of \ac{BNS} merger rates of
$10^{-8}$--$10^{-5}$\,Mpc$^{-3}$\,yr$^{-1}$ \cite{2010CQGra..27q3001A}, or
\ac{NSBH} merger rates of
$6\,\times\,10^{-10}$--$1\,\times\,10^{-6}$\,Mpc$^{-3}$\,yr$^{-1}$
\cite{2010CQGra..27q3001A}. However, the recent work of
\cite{2015MNRAS.448..928K}, which studied known galactic \ac{BNS} systems in
more detail suggest a rate approximately five times lower. The estimates for
\ac{NSBH} systems come entirely from population synthesis models as no such
systems have yet been observed.
%The ranges of rates for \acp{BNS} is obtained from population synthesis models
%together with knowledge of the X known \ac{BNS} systems in our galaxy, whilst
%the range for \acp{NSBH} is entirely from population synthesis as not such
%systems have been observed.

If a single \ac{GW} event is observed in coincidence with a \ac{sGRB} it may be
possible to identify the host galaxy of the \ac{BNS} source.  With this
information a spectroscopic redshift can be very accurately obtained (see
values quoted in e.g. \cite{0004-637X-664-2-1000} and references therein). With
knowledge of the redshift, using current best estimates for the Hubble constant
and other cosmological parameters, the true luminosity distance can be
estimated to $\sim 1\%$ accuracy.  The \ac{GW} measurement acts as a standard
siren also giving us a direct measurement of the luminosity distance to the
source. The accuracy of such a measurement depends on a number of factors
including the accuracy with which the \ac{GW} detector has been calibrated.
Hence, by comparison with the distance estimate from the \ac{sGRB} we can
re-calibrate (or validate) the existing experimentally obtained calibration.

\MP{Move this paragraph to conclusions} For future planned space-based detectors such as eLISA there are several known galactic binary 
systems known as ``verification'' binaries \cite{2006CQGra..23S.809S}, in that their masses and 
orbital parameters allow them to be guaranteed sources for such detectors. These too could be used 
for assessing calibration for such detectors. However, generally for these verification binaries 
the inclination of the system is not known and distances are fairly uncertain, so an accurate 
calibration assessment may not be possible. Although there is at least one (J0651) currently known 
eclipsing white dwarf binary \cite{2041-8205-737-1-L23}, which does provide the system inclination 
\cite{2012ApJ...757L..21H, 2013ASPC..467...47K} and has an approximately 10\% distance uncertainty 
\cite{2041-8205-737-1-L23}, which make it an excellent candidate for an astrophysical calibrator at 
its frequency.

% What do we do in this paper
In this paper we investigate the feasibility of this approach for single coincident 
\ac{GW}--\ac{sGRB} events and establish the validation power of such a calibration technique.  In 
Sec.~\ref{sec:calibration} we briefly summarize the existing experimental technique for \ac{GW} 
detector calibration and its expected accuracy.  We then review the concept of \ac{GW} standard 
sirens and their proposed coincident \ac{EM} signatures, the \acp{sGRB} in
Secs.~\ref{sec:sirens} and~\ref{sec:GRB}. In Secs.~\ref{sec:analysis} and \ref{sec:results} we 
discuss out analysis method and the results, and in Sec.~\ref{sec:discussion} we 
conclude.

@@

The technique used throughout \ac{GW} research in calibrating the detectors is carried out through 
a complicated system of physical manipulations of the Fabry-P\'{e}rot Michelson interferometer; 
where an elaborate feedback system is used to sustain a defined measurement in arm length 
difference between the moving mirrors. A comprehensive description of the calibration procedure (in 
particular for the LIGO detectors during their fifth science run) can be found in 
\cite{2010NIMPA.624..223A}, and a brief description is giving in \cite{Vitale:2012} and references 
therein. For \ac{aLIGO} and \ac{AdV}, at the frequency range used in this project, the error in the 
amplitude calibration is roughly $10\%$ \cite{Vitale:2012}. This is a benchmark set for the error 
estimation using the proposed method in this project.

% Describe our model for the calibration
The \ac{GW} strain is measured through the differential arm length, $\Delta L$, changes of the 
interferometer via $h(f,t) = \Delta L(f,t) / L$, where $L$ is the full arm length. Calibration is 
required to relate the actual measured interferometer error signal output $e(f)$ to $\Delta L$. 
This relation is known as the length response function, $R(f)$, defined such that
\begin{equation}
\Delta L(f,t) = R(f) e(f, t),
\end{equation}
where where we assume $R$ varies only slowly in time (in comparison to transient signal 
time-scales). Calculation of $R$ requires the measurement of various functions within a control 
feedback loop (see \cite{2010NIMPA.624..223A}), which are subject to uncertainties. In this 
study we will assume an estimate of $R$ is available (although in theory we could take on the role 
of estimating $R$ itself), but that it differs from the truth through some unknown scale factor, 
$\scf$, so that at a particular time we have
\begin{equation}\label{eq:scalefactor}
\scf h(f) = h_{\rm m}(f) = \frac{R(f) e(f)}{L},
\end{equation}
where $h(f)$ is the true strain and $h_{\rm m}(f)$ is the measured strain. With this definition it
means that if $\scf > 1$ then a signal would appear to have a larger amplitude (e.g.\ be 
closer) than in reality\footnote{However, the \ac{SNR} would be the correct value as the signal and 
the noise will both contain the scale factor.}, whereas if $\scf < 1$ it would appear to have a 
lower amplitude than reality. In this analysis we will simplify the situation by assuming that 
$\scf$ is a constant with respect to frequency, but in future studies that assumption could be 
dropped and $\scf$ could take some functional form, or piece-wise fit, with respect to $f$ (e.g.\ 
in a similar way to the method of \cite{2013PhRvD..88h4044L} used for fitting differences in 
\ac{PSD} estimates).

@@

% what are sGRBs 
It is believed that \acp{sGRB} (those \acp{GRB} with duration ${<}2$ sec) are
emitted during the merger of
\ac{BNS} or \ac{NSBH} systems ~\cite{1989Natur.340..126E,1992ApJ...395L..83N}. The emission
from these events is highly beamed along the binary rotation axis and hence
only potentially observable for a fraction of the mergers. If the event
exhibits an optical afterglow then the host galaxy can be identified from which
a redshift can be obtained (e.g.~\cite{2005Natur.437..851G}).  The fraction of events with 
associated redshifts is ${\sim}1/3$. The range to which the \emph{Swift} \ac{XRT} has detected
\acp{sGRB} is $z{\sim}2$ and the nearest event with associated redshift is at
$z{\sim}0.1$~\cite{2015GCN..17278...1C} which is approximately equal to the
horizon range of the advanced \ac{GW} detector network for \ac{BNS} systems.

% discuss joint-event rates
The predicted detection rates of \ac{BNS} events (irrespective of an \ac{sGRB}
counterpart) are derived from three main sources, population synthesis models,
extrapolations based on known galactic \ac{BNS} systems, and from observations
of \ac{sGRB} events. Various studies using the first two methods have been
compiled into an overall detection rate of $0.2$--$200$ \ac{BNS} events per
year~\cite{2013arXiv1304.0670L,2010CQGra..27q3001A}\footnote{We note that this
rate is that given in Table 1 of \cite{2013arXiv1304.0670L} for an \ac{aLIGO}
and \ac{AdV} network consisting to two \ac{aLIGO} detectors operating at design
sensitivity. However, a higher rate of $0.4$--$400$ is given in
\cite{2010CQGra..27q3001A}, which may differ due to the use of a third
\ac{aLIGO} detector, but the discrepancy may also have contributions from
slightly different threshold definitions or design sensitivity curves used. We
also note the study mentioned in Sec.~\ref{sec:intro} that potentially lowers
the expected rate by a factor of five.}.  In~\cite{2012MNRAS.425.2668C} using
the rate of observed \acp{sGRB} and assuming a beaming angle of $15^{o}$ a
similar event rate of $1$--$180$ is obtained. The beaming angle is the primary
factor in determining the fraction of detected \ac{GW} \ac{BNS} signals with an
\ac{EM} \ac{sGRB} counterpart and is unfortunately highly uncertain.  Various
authors have provided estimates on the rate of joint detections to be in the
range $0.02$--$7$ per
year~\cite{2013PhRvL.111r1101C,Kelley:2012fl,2014arXiv1405.2254S,2015MNRAS.448.3026W}.

@@

For a non-spinning system the general form of the frequency-domain polarisation
amplitudes are
%
\begin{align}\label{eq:signal} 
\gws_{+}(f) &\propto
\left(1+\cos^{2}\iota\right)D_{L}^{-1} \mathcal{M}^{5/6}f^{-7/6}e^{-i\Psi(f,
\mathcal{M}, t_\mathrm{c}, \phi_{\mathrm{c}})} \nonumber \\ \gws_{\times}(f) &\propto
\left(\cos^{2}\iota\right)D_{L}^{-1}\mathcal{M}^{5/6} f^{-7/6}e^{-i\Psi(f, \mathcal{M},
t_\mathrm{c}, \phi_{\mathrm{c}})} 
\end{align}
%
where the chirp-mass $\mathcal{M}$ is defined as $\mathcal{M}=M\eta^{3/5}$,
with the symmetric mass ratio $\eta=m_{1}m_{2}/M^{2}$.  The total mass
$M=m_{1}+m_{2}$ where $m_{1}$ and $m_{2}$ are the component masses, $\iota$ is
the inclination angle.  The \ac{GW} strain measured
at the $k^{\rm{th}}$ detector is then given by
%
\begin{equation} \label{eq:gravsig} 
\gws^{k}(f) = \big[ F_{+}^{k}(\alpha,
\delta, \psi)\gws_{+}(f) + F_{\times}^k(\alpha, \delta,
\psi)\gws_{\times}(f)\big] 
\end{equation}
%
where $F^{k}_{+}$ and $F^{k}_{\times}$ are the antenna response functions which
are dependent upon the polarisation angle $\psi$ and the sky position of the
source $\alpha$ and $\delta$.

@@

% discuss the inclination and opening angle
Other information relevant for a coincident \ac{sGRB} is that the system is
likely to be relatively close to face-on (and therefore nearly circularly
polarised), which allows us to place prior constraints on the system
inclination angle. Beam opening angles for \acp{sGRB} are relatively poorly
constrained as they are based on only a few jet-break detections. A comparative
study of estimates and lower-limits of opening angles in 13 \acp{sGRB} given
in~\cite{2014ApJ...780..118F} provide a median opening angle value of $\sim
10^{\circ}$. We assume that the beam opening angle is a half-normal
distribution, for which the median is given by $\sqrt{2}\erf{}^{-1}(1/2)\sigma
\sim 0.67\sigma$. It follows that the median of a $\sim 10^{\circ}$ opening
angle corresponds to $\sigma \sim 14.8^{\circ}$. We therefore place a Gaussian
prior on the inclination angle $\iota$ with zero mean and a standard deviation
$\sigma=14.8^{\circ}$.~\MP{Check regarding the face/arse orientation issue.}

@@

% discuss the known sky position
The final important piece of information that we can make use of from the
\ac{sGRB} observation is the sky position of the source. We assume that this is 
known precisely and has been obtained, in conjunction with the source redshift,
from the identification of the host galaxy.

@@

% the scale factor prior
Finally, we require a prior on the calibration scale factors for each detector.
We make the assumption that in general the calibration applied to the detector
data will be correct, so want to use a prior that is peaked at $\scf_k=1$. We
also assume that the prior \ac{PDF} of the calibration scale factor is either
larger or smaller than unity by an equivalent factor, e.g.\ $p(\scf|I)=10$ is
equivalent to $p(\scf|I)=0.1$. For the scale factor in each detector we 
therefore use the following log-normal distribution as our prior
%
\begin{equation}\label{eq:scale_prior}
 p(\scf|I) = \frac{1}{\scf\sigma\sqrt{2\pi}}\exp{\left( -\frac{(\ln{\scf} - \mu)^2}{2\sigma^2} 
\right)},
\end{equation}
%
where we chose $\sigma = 1.07$, which together with the requirement that the
mode is at unity gives $\mu = 1.15$. With these parameters the prior
probability density at 0.1 and 10 is one tenth of that at 1. Note that this
prior does not give a symmetric amount of probability about unity (so our prior
favour \scf\xspace values greater than 1), but given our simulation criterion
described in Section~\ref{sec:simulations} we find that the likelihood
generally overwhelms the prior. As $\scf$ is a scale factor we also have the
further constraint that it must be positive and hence the log-normal prior is a
natural choice.

@@

% Describe the ensemble of simulations used
Separately for the \ac{BNS} and \ac{NSBH} systems we have performed a large
number (several hundred up to $O(1000)$) of simulations at each distance value.
When proposing injections for each simulation the source parameters were
randomly drawn from the prior distributions described in
Section~\ref{sec:priors}, with the sky position being drawn randomly from a
uniform distribution on the sky, the coalescence times held fixed within the
centre of the prior window, and with the calibration scaling factors for each
detector drawn from a Gaussian distribution with a mean of one and standard
deviation of $0.125$ (equivalent to a mean fractional offset of 10\%).  Note
that this distribution is broadly consistent with the scale of calibration
uncertainties that are expected via experimental methods but not the same as
the prior that we assume on $\mathcal{C}$ (see Eq.~\ref{eq:scale_prior}) when
performing our analysis.

% describe the detection cut 
In accepting a proposed injection as one to be analysed we introduced a
criteria that the signal be ``detectable'', based on that used in
\cite{2012PhRvD..85h2002A}. We therefore only analysed injections that had an
\ac{SNR} of $\geq 5.5$ in at least two detectors. Since we assume that these
signals would be coincident with a \ac{sGRB} we do not include the often used
further constraint that the total coherent network \ac{SNR} is greater than 12.
For injections close to the \ac{SNR} threshold, the application of the
threshold resulted in the preferential selection of better oriented systems.
Therefore, at larger distances the population of injected sources was not
uniformly distributed over the sky and instead favoured locations with better
antenna response.  Similarly, despite being already constrained by the prior to
small inclination angles, the accepted injections show preference for sources
very close to circularly polarised.

% final clarification on the simulation distribution
These simulations have allowed us to assess how well on average we would be
able to constrain the calibration scale for a given detected \ac{CBC}-\ac{sGRB}
coincidence.

@@

% mention caveats and future work
The requirement of a coincident \ac{sGRB} with a known distance means that
there will likely be considerable delay (beyond initial \ac{GW} detection) for
the implementation of this method of calibration assessment.  Importantly
however, this method would provide a \MP{semi-independent} consistency check that
the calibration is reasonably accurate.  We acknowledge that we have made a
number of simplifying assumptions, including that the waveforms we use
accurately model the true signal, and that the overall calibration remains
constant over the duration of a signal and its \ac{PSD} estimation. A further
major simplification made here is that the absolute calibration is constant
with frequency.  Other authors (e.g.\ \cite{Vitale:2012,2013PhRvD..88h4044L} have addressed
this issue in the context of robust parameter estimation for astrophysical
sources and in future work we intend to apply related methods for astrophysical
calibration.

@@

>>>>

\begin{abstract}
  We investigate a method to assess the validity of gravitational wave detector calibration 
through the use of gamma-ray bursts as standard
  sirens. Such signals, as measured via gravitational wave
  observations, provide an estimated luminosity distance that is
  subject to uncertainties in the calibration of the data.  If a host
  galaxy is identified for a given source then its redshift can be
  combined with current knowledge of the cosmological
  parameters yielding the true luminosity distance.  This will then allow
  a direct comparison with the estimated value and can validate the accuracy of the original 
calibration. We use simulations of individual detectable gravitational wave signals from binary 
neutron star (BNS) or neutron star-black hole (NSBH) systems, which we assume to be found in 
coincidence with short gamma-ray bursts, to estimate any discrepancy in the overall scaling of the 
calibration for detectors in the Advanced LIGO and Advanced Virgo network. We find that the 
amplitude scaling of the calibration for the LIGO instruments could on average be confirmed to 
within $\sim 10\%$ for a BNS source within 100\,Mpc. This result is largely independent of the 
current detector calibration method and gives an uncertainty that is competitive with that expected 
in the current calibration procedure. Confirmation of the calibration accuracy to within 
$\sim 20\%$ can be found with BNS sources out to $\sim 500$\,Mpc.
\end{abstract}

\pacs{04.80.Nn, 95.55.Ym}
\preprint{\dcc}

@@

% BNS systems are GRBs
It is considered likely that the merger of \ac{BNS} and/or \ac{NSBH} systems,
in addition to emitting detectable \acp{GW}, is also the mechanism for
producing \acp{sGRB} \cite{1992ApJ...395L..83N}. In this scenario these events
produce tightly beamed \ac{EM} emission parallel to the orbital angular
momentum vector of the system. Additional evidence for the coincidence
of \acp{sGRB} with \ac{GW} events are the estimated astrophysical rates of both
phenomena. Observations of \acp{sGRB} give an inferred rate of \ac{sGRB}
producing \ac{CBC} mergers of $8\times 10^{-9}$--$1.1\times
10^{-6}$\,Mpc$^{-3}$\,yr$^{-1}$~\cite{2012MNRAS.425.2668C} which can be
compared to independently obtained estimates of \ac{BNS} merger rates of
$10^{-8}$--$10^{-5}$\,Mpc$^{-3}$\,yr$^{-1}$ \cite{2010CQGra..27q3001A}, or
\ac{NSBH} merger rates of
$6\,\times\,10^{-10}$--$1\,\times\,10^{-6}$\,Mpc$^{-3}$\,yr$^{-1}$
\cite{2010CQGra..27q3001A}\footnote{However, the recent work of
\cite{2015MNRAS.448..928K}, which studied known galactic \ac{BNS} systems in
more detail suggest a rate approximately five times lower.}.
%The estimates for
%\ac{NSBH} systems come entirely from population synthesis models as no such
%systems have yet been observed.
%The ranges of rates for \acp{BNS} is obtained from population synthesis models
%together with knowledge of the X known \ac{BNS} systems in our galaxy, whilst
%the range for \acp{NSBH} is entirely from population synthesis as not such
%systems have been observed.

If a single \ac{GW} event is observed in coincidence with a \ac{sGRB} it may be
possible to identify the host galaxy of the source.  With this
information a spectroscopic redshift can be very accurately obtained (see
values quoted in e.g.\ \cite{0004-637X-664-2-1000} and references therein). With
knowledge of the redshift, using current best estimates for the Hubble constant
and other cosmological parameters, the true luminosity distance can be
estimated to $\sim 1\%$ accuracy.  The \ac{GW} measurement acts as a standard
siren also giving us a direct measurement of the luminosity distance to the
source. The accuracy of such a measurement depends on a number of factors
including the accuracy with which the \ac{GW} detector has been calibrated.
Hence, by comparison with the distance estimate from the \ac{sGRB} we can
re-calibrate (or validate) the existing experimentally obtained calibration.

% What do we do in this paper
In this paper we investigate the feasibility of this approach for single coincident 
\ac{GW}--\ac{sGRB} events and establish the validation power of such a calibration technique.  In 
Sec.~\ref{sec:calibration} we briefly summarize the existing experimental technique for \ac{GW} 
detector calibration and its expected accuracy.  We then review the concept of \ac{GW} standard 
sirens and their proposed coincident \ac{EM} signatures, the \acp{sGRB} in
Secs.~\ref{sec:sirens} and~\ref{sec:GRB}. In Secs.~\ref{sec:analysis} and \ref{sec:results} we 
discuss our analysis method and the results, and in Sec.~\ref{sec:discussion} we 
conclude.

@@

The technique used throughout \ac{GW} research in calibrating the detectors is carried out through 
a complicated system of physical manipulations of the Fabry-P\'{e}rot Michelson interferometer; 
where an elaborate feedback system is used to sustain a defined measurement in arm length 
difference between the moving mirrors. A comprehensive description of the calibration procedure (in 
particular for the LIGO detectors during their fifth science run) can be found in 
\cite{2010NIMPA.624..223A}, and a brief description is giving in \cite{Vitale:2012} and references 
therein. For \ac{aLIGO} and \ac{AdV}, at the frequency range used in this project, the error in the 
amplitude calibration is expected to be roughly $10\%$ \cite{Vitale:2012}. This is a benchmark set 
for the error estimation using the proposed method in this project.

% Describe our model for the calibration
The \ac{GW} strain is measured through the differential arm length, $\Delta L$, changes of the 
interferometer via $h(f,t) = \Delta L(f,t) / L$, where $L$ is the full arm length. Calibration is 
required to relate the actual measured interferometer error signal output $e(f)$ to $\Delta L$. 
This relation is known as the length response function, $R(f)$, defined such that
\begin{equation}
\Delta L(f,t) = R(f) e(f, t),
\end{equation}
where we assume $R$ varies only slowly in time (in comparison to transient signal 
time-scales). Calculation of $R$ requires the measurement of various functions within a control 
feedback loop (see \cite{2010NIMPA.624..223A}), which are subject to uncertainties. In this 
study we will assume an estimate of $R$ is available (although in theory we could take on the role 
of estimating $R$ itself), but that it differs from the truth through some unknown scale factor, 
$\scf$, so that at a particular time we have
\begin{equation}\label{eq:scalefactor}
\scf h(f) = h_{\rm m}(f) = \frac{R(f) e(f)}{L},
\end{equation}
where $h(f)$ is the true strain and $h_{\rm m}(f)$ is the measured strain. With this definition it
means that if $\scf > 1$ then a signal would appear to have a larger amplitude (e.g.\ be 
closer) than in reality\footnote{However, the \ac{SNR} would be the correct value as the signal and 
the noise will both contain the scale factor.}, whereas if $\scf < 1$ it would appear to have a 
lower amplitude than reality. In this analysis we will simplify the situation by assuming that 
$\scf$ is a constant with respect to frequency, but in future studies that assumption could be 
dropped and $\scf$ could take some functional form, or piece-wise fit, with respect to $f$ (e.g.\ 
in a similar way to the method of \cite{2013PhRvD..88h4044L} used for fitting differences in 
\ac{PSD} estimates).

@@

% what are sGRBs 
It is believed that \acp{sGRB} (those gamma-ray bursts with duration ${<}2$ sec) are
emitted during the merger of
\ac{BNS} or \ac{NSBH} systems ~\cite{1989Natur.340..126E,1992ApJ...395L..83N}. The emission
from these events is highly beamed along the binary rotation axis and hence
only potentially observable for a fraction of the mergers. If the event
exhibits an optical afterglow then the host galaxy can be identified from which
a redshift can be obtained (e.g.~\cite{2005Natur.437..851G}).  The fraction of events with 
associated redshifts is ${\sim}1/3$. The range to which the \emph{Swift} \ac{XRT} has detected
\acp{sGRB} is $z{\sim}2$ and the nearest event with associated redshift is at
$z{\sim}0.1$~\cite{2015GCN..17278...1C} which is approximately equal to the
horizon range of the advanced \ac{GW} detector network for \ac{BNS} systems.

% discuss joint-event rates
The predicted detection rates of \ac{BNS} events (irrespective of an \ac{sGRB}
counterpart) are derived from three main sources, population synthesis models,
extrapolations based on known galactic \ac{BNS} systems, and from observations
of \ac{sGRB} events. Various studies using the first two methods have been
compiled into an overall detection rate of $0.2$--$200$ \ac{BNS} events per
year~\cite{2013arXiv1304.0670L,2010CQGra..27q3001A}\footnote{We note that this
rate is that given in Table 1 of \cite{2013arXiv1304.0670L} for an \ac{aLIGO}
and \ac{AdV} network consisting to two \ac{aLIGO} detectors operating at design
sensitivity. However, a higher rate of $0.4$--$400$ is given in
\cite{2010CQGra..27q3001A}, which may differ due to the use of a third
\ac{aLIGO} detector, but the discrepancy may also have contributions from
slightly different threshold definitions or design sensitivity curves used. We
also note the study mentioned in Sec.~\ref{sec:intro} that potentially lowers
the expected rate by a factor of five.}.  In~\cite{2012MNRAS.425.2668C} using
the rate of observed \acp{sGRB} and assuming a beaming angle of $15^{\circ}$ a
similar event rate of $1$--$180$ is obtained. The beaming angle is the primary
factor in determining the fraction of detected \ac{GW} \ac{BNS} signals with an
\ac{EM} \ac{sGRB} counterpart and is unfortunately highly uncertain.  Various
authors have provided estimates on the rate of joint detections to be in the
range $0.02$--$7$ per
year~\cite{2013PhRvL.111r1101C,Kelley:2012fl,2014arXiv1405.2254S,2015MNRAS.448.3026W}.

@@

For a non-spinning system the general form of the frequency-domain polarisation
amplitudes is
%
\begin{align}\label{eq:signal} 
\gws_{+}(f) &\propto
\left(1+\cos^{2}\iota\right)D_{L}^{-1} \mathcal{M}^{5/6}f^{-7/6}e^{-i\Psi(f,
\mathcal{M}, t_\mathrm{c}, \phi_{\mathrm{c}})}, \nonumber \\
\gws_{\times}(f) &\propto
\left(\cos^{2}\iota\right)D_{L}^{-1}\mathcal{M}^{5/6} f^{-7/6}e^{-i\Psi(f, \mathcal{M},
t_\mathrm{c}, \phi_{\mathrm{c}})} 
\end{align}
%
where the chirp-mass $\mathcal{M}$ is defined as $\mathcal{M}=M\eta^{3/5}$,
with the symmetric mass ratio $\eta=m_{1}m_{2}/M^{2}$, total mass
$M=m_{1}+m_{2}$ (where $m_{1}$ and $m_{2}$ are the component masses), and $\iota$ is
the inclination angle.  The Fourier transform of the \ac{GW} strain measured
at the $k^{\rm{th}}$ detector is then given by
%
\begin{equation} \label{eq:gravsig} 
\gws^{k}(f) = \big[ F_{+}^{k}(\alpha,
\delta, \psi)\gws_{+}(f) + F_{\times}^k(\alpha, \delta,
\psi)\gws_{\times}(f)\big] 
\end{equation}
%
where $F^{k}_{+}$ and $F^{k}_{\times}$ are the antenna response functions which
are dependent upon the polarisation angle $\psi$ and the sky position of the
source $\alpha$ and $\delta$.

@@

% discuss the inclination and opening angle
Other information relevant for a coincident \ac{sGRB} is that the system is
likely to be relatively close to face-on (and therefore nearly circularly
polarised), which allows us to place prior constraints on the system
inclination angle. Beam opening angles for \acp{sGRB} are relatively poorly
constrained as they are based on only a few jet-break detections. A comparative
study of estimates and lower-limits of opening angles for 13 \acp{sGRB} given
in~\cite{2014ApJ...780..118F} provides a median opening angle value of $\sim
10^{\circ}$. We assume that the beam opening angle is a half-normal
distribution, for which the median is given by $\sqrt{2}\erf{}^{-1}(1/2)\sigma
\sim 0.67\sigma$. It follows that the median of a $\sim 10^{\circ}$ opening
angle corresponds to $\sigma \sim 14.8^{\circ}$. We therefore place a Gaussian
prior on the inclination angle $\iota$ with zero mean and a standard deviation
$\sigma=14.8^{\circ}$. This, and the choice of polarisation angle prior range given below, 
explicitly fixes the handedness of the system's rotation (or alternatively whether it is face-on or 
back-on). If we were to allow the inclination to have two modes (one for face-on and one 
for back-on), or alternatively to double the range of the polarisation angle, it would have no 
effect on our ability to estimate the signal amplitude as the two modes should be identical with 
regards to the amplitude parameter probability distributions. However, we are consistent between the 
range of simulated signals discussed in Sec.~\ref{sec:simulations} and the initial ranges stated.

@@

% discuss the known sky position
The final important piece of information that we can make use of from the
\ac{sGRB} observation is the sky position of the source. We assume that this is 
known precisely and has been obtained, in conjunction with the source redshift,
from the identification of the host galaxy. This fixes the time delay between detectors to a known 
value and, along with the polarisation angle, defines the antenna response patterns.

@@

% the scale factor prior
Finally, we require a prior on the calibration scale factors for each detector.
We make the assumption that in general the calibration applied to the detector
data will be close to being correct, so want to use a prior that is peaked at $\scf_k=1$. We
also assume that the prior \ac{PDF} of the calibration scale factor is either
larger or smaller than unity by an equivalent factor, e.g.\ $p(\scf|I)=10$ is
equivalent to $p(\scf|I)=0.1$. For the scale factor in each detector we 
therefore use the following log-normal distribution as our prior
%
\begin{equation}\label{eq:scale_prior}
 p(\scf|I) = \frac{1}{\scf\sigma\sqrt{2\pi}}\exp{\left( -\frac{(\ln{\scf} - \mu)^2}{2\sigma^2} 
\right)},
\end{equation}
%
where we chose $\sigma = 1.07$, which together with the requirement that the
mode is at unity gives $\mu = 1.15$. With these parameters the prior
probability density at 0.1 and 10 is one tenth of that at 1. Note that this
prior does not give a symmetric amount of probability about unity (so our prior
favours \scf\xspace values greater than 1), but given our simulation criterion
described in Section~\ref{sec:simulations} we find that the likelihood
generally overwhelms the prior. As $\scf$ is a scale factor we also have the
further constraint that it must be positive and hence the log-normal prior is a
natural choice.

@@

% Describe the ensemble of simulations used
Separately for the \ac{BNS} and \ac{NSBH} systems we have performed a large
number (several hundred up to $O(1000)$) of simulations at each distance value.
When proposing injections for each simulation the source parameters were
randomly drawn from the prior distributions described in
Section~\ref{sec:priors}, with the sky position being drawn randomly from a
uniform distribution on the sky, the coalescence times held fixed within the
centre of the prior window, and with the calibration scaling factors for each
detector drawn from a Gaussian distribution with a mean of one and standard
deviation of $0.125$ (equivalent to a mean fractional offset of 10\%).  Note
that this distribution is broadly consistent with the scale of calibration
uncertainties that are expected via experimental methods, but is not the same as
the prior that we assume on $\mathcal{C}$ (see Eq.~\ref{eq:scale_prior}) when recovering 
signals.

% describe the detection cut 
In accepting a proposed injection as one to be analysed we introduced a
criteria that the signal be ``detectable'', based on that used in
\cite{2012PhRvD..85h2002A}. We therefore only analysed injections that had an
\ac{SNR} of $\geq 5.5$ in at least two detectors. Since we assume that these
signals would be coincident with a \ac{sGRB} we do not include the often used
further constraint that the total coherent network \ac{SNR} is greater than 12.
For injections close to the \ac{SNR} threshold, the application of the
threshold resulted in the preferential selection of better oriented systems.
Therefore, at larger distances the population of injected sources was not
uniformly distributed over the sky and instead favoured locations with better
antenna response.  Similarly, despite already being constrained by the prior to
small inclination angles, the accepted injections show preference for sources
very close to circularly polarised.

% final clarification on the simulation distribution
These simulations have allowed us to assess how well on average we would be
able to constrain the calibration scale for a detected \ac{CBC}-\ac{sGRB}
coincidence.

@@

% mention caveats and future work
The requirement of a coincident \ac{sGRB} with a known distance means that
there will likely be considerable delay (beyond initial \ac{GW} detection) for
the implementation of this method of calibration assessment.  Importantly
however, this method would provide an independent\footnote{There are caveats to this independence. 
We rely on a source having been detected, which most likely implies that calibration is not grossly 
inaccurate and is fairly consistent between detectors.} consistency check that the calibration is 
reasonably accurate.  We acknowledge that we have made a number of simplifying assumptions, 
including that the waveforms we use accurately model the true signal, and that the overall 
calibration remains constant over the duration of a signal and its \ac{PSD} estimation. A further
major simplification made here is that the absolute calibration is constant
with frequency.  Other authors (e.g.\ \cite{Vitale:2012,2013PhRvD..88h4044L}) have addressed
this issue in the context of robust parameter estimation for astrophysical
sources and in future work we intend to apply related methods for astrophysical
calibration.

@@

It is also worth noting that astrophysical calibration may be possible for future planned 
space-based detectors such as eLISA \cite{2013GWN.....6....4A}. For such detectors there are 
several 
galactic binary systems known as ``verification'' binaries \cite{2006CQGra..23S.809S}, in that 
their masses and orbital parameters allow them to be guaranteed sources. These too could be used 
for assessing detector calibration. However, generally for these verification binaries 
the inclination of the system is not known and distances are fairly uncertain, so a precise 
calibration assessment may not be possible. Although there is at least one currently known 
eclipsing white dwarf binary (J0651) \cite{2041-8205-737-1-L23}, which does provide the system 
inclination \cite{2012ApJ...757L..21H, 2013ASPC..467...47K} and has an approximately 10\% distance 
uncertainty \cite{2041-8205-737-1-L23}, which makes it an excellent candidate for an astrophysical 
calibrator.
